---
title: "Final Paper"
author: "Sophia Angleton, Fiona Debernardi, Stephen Anti"
date: "2024-12-11"
output:
  html_document:
    df_print: paged
---
```{r}
#| include: FALSE
library(tidyverse)
library(janitor)
library(tidyr)
library(readr)
library(rio)
library(knitr)
library(kableExtra)
library(dplyr)
library(psych) #generate metrix w scatterplot and cor
#import dataset
AlternateRuns <- read_csv("AlternateRuns.csv")
```

## Abstract

This study is interested in understanding the relationship between task-switching and the cognitive stability-flexibility tradeoff. The design of this study will involve task-switching in an alternating runs manner, where there are two cues this task switches between, shape and color and these alternate in cycles of 4, each run being cycles of 4. This study will use alternating runs task-switching to understand the relationship between cognitive stability and cognitive flexibility. We will use the stability-flexibility tradeoff to inform our predictions, where we predict that 1) There is a negative relationship between expression of stability and expression of flexibility. We will see this relationship on three levels, within-subjects, in individual differences, and on the experimental level. Through a basic analysis of data, we saw there was a difference between stability and flexibility through a difference in switch cost measured in response time. This shows that there is a presence of stability and flexibility informing our task-switching study, however further analysis is required to understand the directionality of the relationship. 

## Introduction

The paradigm between flexibility and cognitive stability is a widely known and established relationship in the world of cognitive science. Both cognitive stability and flexibility are aspects of executive functioning, or what we understand as self-control (Merian, 2010). Cognitive flexibility centers itself in a domain of executive functioning called mental set shifting, and in previous literature, has been understood through the use of the task-switching paradigm. Tasks such as odds-evens, the Stroop task, or other alternating tasks where through the use of cues, participants need to identify an alternating attribute of a stimuli (Merian, 2010). For this study, the task-switch paradigm is constructed through the identification of one of two attributes, color or shape, where they identify color of the stimulus or shape of the stimulus, respectively. According to the established stability/flexibility tradeoff, when one is performing a task that requires more cognitive stability, it is harder to also be more cognitively flexible. This is seen in the task-switch paradigm within-subjects where they have less switch-costs when not switching task cues compared to higher switch-costs when switching from one cue to another (Goschke, 2000). 

However, a recent reevaluation of the generalizability of the stability/flexibility tradeoff has posited that tradeoffs originally thought to explain a plethora of cognitive models, occur only in highly specified contexts (Mayr & Graetz). Instead, there is newfound evidence of an anti-tradeoff pattern, meaning there is co-occurrence of cognitive stability and flexibility depending on the level of resolution encoding (Mayr et al.). These recent findings in the field of decision-making suggest that the stability-flexibility trade off may not be as strong as once thought, however in this study we are still predicting we will see a negative relationship between the switch (flexibility) and no-switch (stability) variables through a comparison of error rate and response time until further studies explain more about a potential anti-tradeoff occurring. Indeed, using the stability-flexibility tradeoff to inform our predictions, we predict that: 1) There is a negative relationship between expression of stability and expression of flexibility. We will see this relationship on three levels, within-subjects, in individual differences, and on the experimental level. We will calculate the difference in reaction times between no-switch and switch trials. A smaller reaction time difference is interpreted as a higher level of stability, whereas a longer reaction time indicates a higher level of flexibility (Goschke, 2000). 

## Methods

```{r}
#rename columns to understand better
alt_run <- AlternateRuns %>% 
  rename(dimshape = dim1, dimcolor = dim2, RT = time, correct = cor, response = res)
alt_run
```
```{r}
#lets look at the top RTs. 
alt_run %>% 
  arrange(desc(RT))

#WHAT TO DO: z-score on each seq position x switch x ambiguity on RT then z-score on each block (to account for some participants only doing 7 instead of 8 blocks) 

#STEP 1: separate switch trials, c(1,5) and control trials !c(1,5)


alt_run <- alt_run %>% 
  mutate(trial_type = if_else(cycle %in% c(1,5), 'Switch', 'No-Switch')) #saying if its 1 or 5 assign switch, else, assign control. so cool. need to also throw out the first trial after each block bc its not a switch or noswitch. need to separate into cycle positions of 4, so total conditions are 16 
alt_run

#STEP 2: Look at mean RTs by Trial_type, look at z-scores

z_scoretrial <- alt_run %>% 
  group_by(trial_type) %>% 
  mutate(z= (RT- mean(RT))/sd(RT)) %>% 
  summarize(z_score_mean = mean(z, na.rm = TRUE),z_score_sd = sd(z, na.rm = TRUE), response_time_mean = mean(RT, na.rm = TRUE),response_time_mean = sd(RT, na.rm= TRUE))

z_scoretrial

#STEP 2.1: Interpret the data

#so this is telling us that our mean z-score for both control and switch is basically 0 (which is what we want to see) and that our z sd is 1 (which is also what we want to see). looking at the mean for RT in both switch and control, we see that the response time means tend to be a lot longer on average than the average response time for control trials (non-switch trials). this is so cool!

rtdif <- z_scoretrial %>% 
  summarize(meandif = 889.9560 - 579.1123) 
rtdif #difference in means by 310.8437 where the switch trial takes 310.84 ms longer than the control or non-switch trials. 

# Identify outliers (z-scores beyond -3 and 3)

outliers_IRT <- alt_run %>%
  group_by(trial_type) %>% 
  mutate(z_score = (RT- mean(RT))/sd(RT)) %>% 
  filter(abs(z_score) > 3)
outliers_IRT

# Calculate the mean and standard error of RT for each trial_type
summary_data <- alt_run %>% 
  group_by(trial_type) %>% 
  summarise( mean_RT = mean(RT), se_RT = sd(RT) / sqrt(n()) ) %>% mutate(lower_CI = mean_RT - qt(0.975, df = n() - 1) * se_RT, upper_CI = mean_RT + qt(0.975, df = n() - 1) * se_RT)
summary_data

```

## Results

First, we created a density plot to examine how response times (RT) were distributed in the data set (See Figure 1). We wanted to first examine how participants were responding across all conditions, so we averaged participants response times and then put them into this density plot. The average response time (as indicated by the red dashed line) was 860.44 ms and the standard deviation was 697.28 ms.


```{r echo=FALSE}
mean_rt <- mean(alt_run$RT, na.rm = TRUE)
mean_rt
sd_rt <-  sd(alt_run$RT, na.rm = TRUE)

alt_run %>% 
  ggplot(aes(x=RT)) +
  geom_histogram(aes(y = after_stat(density)), fill = 'darkgreen', color = 'darkblue') +
  geom_vline(aes(xintercept = mean_rt) , color = 'red', linetype = 'dashed', size = 1.5) +
  theme_minimal() +
  stat_function(fun = dnorm, args = list(mean = mean_rt, sd = sd_rt), col = 'gold', size = 1.5) +
  labs(x = 'Response Times (ms)', y = 'Density', title = 'Density plot of Response Times', subtitle = 'The mean and normal density curve of RTs')
```

Next, we examined whether the participants' response times in the control condition varied from their response times in the switch condition. We found that average response time varied by condition. In the control condition, participants responded more quickly (M = 747.59, SD = 580.69). In the switch condition, participants had slower response times (M = 1186.02, SD = 891.09). We pivoted the data to create separate columns for each condition, and then we calculated the mean and standard deviation of each column. We then created a table to depict these descriptives.

```{r}
alt_run_wide <- alt_run %>%
  pivot_wider(
    names_from = trial_type,          # Use trial_type as column names
    values_from = RT,      # Fill columns with response_time values
    names_prefix = "response_time_"   
  )

# Convert the relevant columns to numeric for the descriptives table
alt_run_wide$`response_time_No-Switch` <- as.numeric(as.character(alt_run_wide$`response_time_No-Switch`))
alt_run_wide$`response_time_Switch` <- as.numeric(as.character(alt_run_wide$`response_time_Switch`))

# Descriptive statistics for the control trial (No-Switch)
mean_rt_control <- mean(alt_run_wide$`response_time_No-Switch`, na.rm = TRUE)
median_rt_control <- median(alt_run_wide$`response_time_No-Switch`, na.rm = TRUE)
sd_rt_control <- sd(alt_run_wide$`response_time_No-Switch`, na.rm = TRUE)

# Descriptive statistics for the switch trial
mean_rt_switch <- mean(alt_run_wide$`response_time_Switch`, na.rm = TRUE)
median_rt_switch <- median(alt_run_wide$`response_time_Switch`, na.rm = TRUE)
sd_rt_switch <- sd(alt_run_wide$`response_time_Switch`, na.rm = TRUE)

summary_table <- data.frame(
  Statistic = c("Mean RT", "Median RT", "Standard Deviation"),
  `No-Switch (Control)` = c(mean_rt_control, median_rt_control, sd_rt_control),
  `Switch` = c(mean_rt_switch, median_rt_switch, sd_rt_switch)
)

# Round the numeric columns
summary_table <- summary_table %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) 

# Display the table using kable (without LaTeX or interactive features)
library(knitr)

# Use kable to display the table
kable(summary_table, caption = "Descriptive Statistics for Response Times")
```
Finally, we created a bar chart to show the difference in response times by condition. This chart is another way of visualizing the results that we found above. The chart shows that when participants were told by the instructions to switch from one task to another (switch trial), they took longer than when they were doing the same task repeatedly (control trial).

```{r}
#lets look at the top RTs. 
alt_run %>% 
  arrange(desc(RT))

#WHAT TO DO: z-score on each seq position x switch x ambiguity on RT then z-score on each block (to account for some participants only doing 7 instead of 8 blocks) 

#STEP 1: separate switch trials, c(1,5) and control trials !c(1,5)


alt_run <- alt_run %>% 
  mutate(trial_type = if_else(cycle %in% c(1,5), 'Switch', 'No-Switch')) #saying if its 1 or 5 assign switch, else, assign control. so cool. need to also throw out the first trial after each block bc its not a switch or noswitch. need to separate into cycle positions of 4, so total conditions are 16 
alt_run

#STEP 2: Look at mean RTs by Trial_type, look at z-scores

z_scoretrial <- alt_run %>% 
  group_by(trial_type) %>% 
  mutate(z= (RT- mean(RT))/sd(RT)) %>% 
  summarize(z_score_mean = mean(z, na.rm = TRUE),z_score_sd = sd(z, na.rm = TRUE), response_time_mean = mean(RT, na.rm = TRUE),response_time_mean = sd(RT, na.rm= TRUE))

z_scoretrial

#STEP 2.1: Interpret the data

#so this is telling us that our mean z-score for both control and switch is basically 0 (which is what we want to see) and that our z sd is 1 (which is also what we want to see). looking at the mean for RT in both switch and control, we see that the response time means tend to be a lot longer on average than the average response time for control trials (non-switch trials). this is so cool!

rtdif <- z_scoretrial %>% 
  summarize(meandif = 889.9560 - 579.1123) 
rtdif #difference in means by 310.8437 where the switch trial takes 310.84 ms longer than the control or non-switch trials. 

# Identify outliers (z-scores beyond -3 and 3)

outliers_IRT <- alt_run %>%
  group_by(trial_type) %>% 
  mutate(z_score = (RT- mean(RT))/sd(RT)) %>% 
  filter(abs(z_score) > 3)
outliers_IRT

# Calculate the mean and standard error of RT for each trial_type
summary_data <- alt_run %>% 
  group_by(trial_type) %>% 
  summarise( mean_RT = mean(RT), se_RT = sd(RT) / sqrt(n()) ) %>% mutate(lower_CI = mean_RT - qt(0.975, df = n() - 1) * se_RT, upper_CI = mean_RT + qt(0.975, df = n() - 1) * se_RT)
summary_data

#segregate based on cycle position and add 95% CI

cycle_plot <- summary_data %>% 
  ggplot(aes(trial_type, mean_RT)) +
  geom_col(fill= '#9CAF88') +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = 0.2, color = 'brown') +
  labs(title = 'Trial Type By Response Time', subtitle = 'Relationship between response times and Switch/No-Switch trials', x= "Trial Type", y= 'Response Time (ms)')+
  theme_minimal() 
cycle_plot
  
```

##  Discussion

The main goal of this study was to investigate task-switching and the trade off between cognitive stability and flexibility. The overall takeaway is that when participants need to switch instructions, their reaction times to the tasks are slower compared to when they are doing the same task repeatedly. In the future, more research is needed to examine whether the differences in responses to the conditions are statistically significant.


\newpage

## References